A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2017-01-31 rigid_bodies.tex
nve review

\section{New Theory}
Exact solution for torque-free rotation
In a previous paper\cite{Abreu_2017} we presented a formulation for the motion of rigid bodies which was simplified by the introduction of a particular factorization of the rotational matrix. 
quaternion $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ and satisfy the relation\cite{Goldstein2002}
\begin{equation}
\label{eq:norm_eq_1}
\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1.
\end{equation}

Reversely, the matrix $\mt A$ can be directly computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray}.
\end{equation*}

Differentiating Eq.~\ref{eq:norm_eq_1} with respect to time yields
\begin{equation}
\label{eq:diff_qTq}
\tr{\vt q}\dot{\vt q} = 0,
\end{equation}
which implies that, for Eq.~\ref{eq:norm_eq_1} to be preserved upon an infinitesimal motion, the rate of change $\dot{\vt q}$ must lie in the hyperplane orthogonal to $\vt q$.

Let us now introduce two new matrix-valued functions that depend on the quaternion components as
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}

The most remarkable feature is that
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation}

The equation above is a simple and convenient factorization of matrix $\mt A$, since the entries of both $\mt B$ and $\mt C$ depend linearly on the Euler parameters. As demonstrated in Appendix~\ref{sec:auxiliary_math}, this factorization provides a relation between the angular velocity vector $\vt \omega$ and the rates of change of the quaternion components, which is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}

Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$.

\section{Introduction}
\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}
\cite{Omelyan_2011} multiple time scale md for fluids with orientational dof. canonical and isokinetic ensembles.
Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations. This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further.

One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach. This is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully employed in plain MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so forth. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

In this paper, we present a simplified formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases. Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom. We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions. For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.

The paper is organized as follows. In Sec.~\ref{sec:mechanics}, we present a mathematical reformulation of the Hamiltonian equations of motion for rigid bodies in terms of unit quaternions. In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVE and NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Moreover, we present results related to kinetic energy partition in the NVE case, analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

Throughout this paper, we use the word ``quaternion'' as a synonym for a four-dimensional vector, avoiding other interpretations. Thus, only ordinary matrix/vector operations are present in the formulation. Some mathematical details are given in the Appendix.

\section{Mechanics of Rigid Body Motion}
\label{sec:mechanics}

In this section, we first review certain aspects of the kinematics of rigid bodies which are relevant to the present study. Unit quaternions are employed to represent the orientations of rigid bodies. Then, we introduce the use of a particular factorization of the rotation matrix, from which a mathematical relation arises between the rate of change of a four-dimensional quaternion and a three-dimensional angular velocity vector. An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation.

\subsection{Kinematics}

\subsection{Hamiltonian Dynamics}
\label{sec:hamiltonian}

The relation between $\vt q$ and $\vt \omega$ (Eq.~\ref{eq:relation_qdot_omega}) is used here to develop a Hamiltonian formalism for the motion of a rigid body. As usual, it can be described by detaching translation and rotation, represented by ${\vt r}(t)$ and ${\vt q}(t)$, respectively. The position of each particle $j$ will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$. we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \label{eq:EDO_q} \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:EDO_pi}
\end{align}
\end{subequations}

Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body can be rewritten as
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\end{equation}

Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi}:
\begin{subequations}
\label{eq:EDO_system_alternative}
\begin{align}
&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
\end{align}
\end{subequations}

\subsection{Symplectic Numerical Integration}
\label{sec:symplectic}

An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element. 
\subsection{Exact Solution for Free Rotation}
\label{sec:free_rotation}

\section{Molecular Dynamics}
\label{sec:molecular_dynamics}

In this section, we present factorization schemes for generating molecular dynamics algorithms in the microcanonical (NVE) and canonical (NVT) ensembles. In both cases, a three-dimensional system is considered as consisting of $N$ interacting, non-collinear rigid bodies occupying a fixed volume $V$. An index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, including its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$). Thus, the phase space of the whole system now contains $3N$ spatial coordinates and $4N$ quaternion components, along with their $7N$ conjugate momenta. Of course, $2N$ implicit constraints are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.

\subsection{Microcanonical Ensemble}

Considering the system described above, with $\vt r$ and $\vt q$ now representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
\begin{equation}
\label{eq:H_NVE}
\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
\end{equation}

As usual, sampling of a microcanonical ensemble can be carried out by a Hamiltonian dynamics in phase space, which will have $\mathcal{H} = E$ as an integral of motion. The dynamics of this system is governed by a Liouville operator expressed as in Eq.~\ref{eq:full_operator}, but now with individual operators defined as
\[
\begin{split}
i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right).
\end{split}
\]

Therefore, a symplectic integrator can be devised using the same splitting scheme of Sec.~\ref{sec:symplectic}, with the action of each propagator (boost, free translation, and free uniaxial rotation) occurring simultaneously for all rigid bodies in the system. If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\sum_{i=1}^N \vt p_i = \vt P$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\sum_{i=1}^N(\vt p_i t - m_i \vt r_i) = \boldsymbol{\mathcal G}$. If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.

>>>>

\section{Introduction}
\cite{Omelyan_2011} multiple time scale md for fluids with orientational dof. canonical and isokinetic ensembles.
systematic errors
\begin{itemize}
\item[1] Kamberaj\cite{Kamberaj2005}\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}
assumption equipartition of the kinetic energy between the translational and rotational degrees of freedom. We demonstrate that this assumption is not valid and it dramatically affects the drift.
to our knowledge, have not been reported in the
order of the operators
\item[2] our paper review
\item[3] coarse grained: 
Along with advances in software and hardware, improvements in molecular dynamics (MD) algorithms have allowed researchers to extend both the time and the length scales attainable in atomistic simulations. This progress has fomented the ambitious attempt of elucidating complex physical phenomena such as the mechanisms underlying molecular motors, pathways for nanoparticle assembly, and interfacial phenomena, to name just a few. Because of its complexity and large computational demand, such a challenging task requires strategies to enlarge the achievable scales even further. One of the strategies employed to speed up MD simulations is the suppression of unimportant degrees of freedom, as might be the case of intramolecular vibrations. Iterative methods such as SHAKE\cite{Ryckaert1977} and RATTLE\cite{Andersen1983} achieved this suppression by adding constraints to the equations of motion. This is the standard approach in the common case of molecular models with fixed bond lengths and bending angles, but flexible dihedral angles. However, for molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach. This is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble. Being symplectic implies phase-space volume preservation and time reversibility, which are sufficient conditions for numerical integrators to be used in hybrid Monte Carlo (HMC) methods.\cite{Duane1987} In this context, Miller \textit{et al}.\cite{Miller2002} have introduced a symplectic integrator for the rotational motion represented in terms of unit quaternions. This approach has been successfully employed in plain MD as well as in HMC methods to simulate a wide variety of systems composed of rigid molecules, including liquid water,\cite{Sakamaki2011, Reinhardt2012, Palmer2014, Gonzales2014} ice,\cite{Geiger2014} hydrates,\cite{Tribello2009, Gorman2012} gases diffusing in metal-organic frameworks,\cite{Ghoufi2010} and so forth. It has also been applied for molecules designed as collections of interconnected rigid bodies, which is a coarse-graining strategy used for proteins,\cite{Terada2003}, molecular machines,\cite{Akimov2008, Konyukhov2010} nanoparticles,\cite{Knorowski2012, Patra2013} etc.

\end{itemize}

In this paper, we report formulation for the motion of rigid bodies in the microcanonical (NVE) and canonical (NVT) ensembles, as well as numerical integrators for both cases. Our contribution in the NVE case resides on a factorization of the rotation matrix that allows a Hamiltonian formalism to be devised without the need of extending the rotational phase space, thus simplifying the formulation of Miller \textit{et al}.\cite{Miller2002} In the NVT case, both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which two thermostat chains act separately on the rotational and the translational degrees of freedom. We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions. For this, we derive two alternative expressions: one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.

The paper is organized as follows. In Sec.~\ref{sec:molecular_dynamics}, we devise numerical integrators for both the NVT dynamics, whose performance and correctness we assess in Sec.~\ref{sec:numerical_results}. Moreover, we analyze the statistical distributions of the sampled energies in the NVT case, and check the equation we developed for computing pressure. Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}.

\subsection{Microcanonical Ensemble: A Review}
algebraic manipulations
Prompted by the introduction of a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble.
Nos\'{e}-Hoover chain thermostat. the non-symplectic character, we used the approximate solution for rotation, as we high quality and non-Hamiltonian dynamics admitos no backward error analyisis. We now will briefly review the key expressions of the metioned paper. First, recall that the translational and rotational degrees of freedom are decoupled. Mathematicaly, the position of a particle $j$,  constituent of a rigid body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the position of the center of mass of the body, $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix and $\vt d_j$ is the position of the particle in the body-fixed frame of reference. Note that unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ are employed to represent the orientations of rigid bodies, which satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\vt q} = 0$.
The matrix $\mt A$ is computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray},
\end{equation*}
which admits the following factorization:
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation} 
that depend on the quaternion components as
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}
Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} aconvenient factorization of matrix $\mt A$. This factorization provides a relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rates of change of the quaternion components, which is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}
It avoids the use of an ficticious angular velocity component, as Miller~\textit{et al.}\cite{Miller2002}. An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation. Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$.

For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), and kinetic energy as follows:

\begin{equation}
\label{eq:H_NVE}
\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right),
\end{equation}

\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\end{equation}

where an index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, with $m_i$ is the mass and  $\mt I_i$ the inertia tensor which in the principal frame is a diagonal matrix withe entries given by ($I_1$,$I_2$,$I_3$). The equations of motion 
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau \label{eq:edo_pi}
\end{align}
\end{subequations}

consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math} of the aforementione peper.

%Finally, we open the products $\mt B \vt \omega$ and $\mt \Omega \vt \omega$ as in Eq.~\ref{eq:product_B_vector}, which leads us to the following new form for the equations of motion in Eqs.~\ref{eq:EDO_q} and \ref{eq:EDO_pi}:
%\begin{subequations}
%\label{eq:EDO_system_alternative}
%\begin{align}
%&\dot{\vt q} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt q \label{eq:edo_q} \\
%&\dot{\vt \pi} = \frac{1}{2} \sum_{k=1}^3 \omega_k \hat{\mt B}_k \vt \pi + 2 \mt C \vt \tau %\label{eq:edo_pi}
%\end{align}
%\end{subequations}

The phase space of the whole system contains $3N$ spatial coordinates and $4N$ quaternion components, along with their $7N$ conjugate momenta. Of course, $2N$ implicit constraints are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$. A Hamiltonian dynamics in phase space will have $\mathcal{H} = E$ as an integral of motion. The dynamics of this system is governed by a Liouville operator with individual operators defined as
\[
\begin{split}
i\!L_b &= \sum_{i=1}^N \left( \tr{\vt F}_i \diff{}{\vt p_i} + 2 \tr{\vt \tau}_i \tr{\mt C}_i \diff{}{\vt \pi_i} \right), \\
i\!L_t &= \sum_{i=1}^N \frac{\tr{\vt p}_i}{m_i}\diff{}{\vt r_i}, \quad \text{and} \\
i\!L_k &= \frac{1}{2} \sum_{i=1}^N {\omega_i}_k \left( \tr{\vt q}_i\tr{\hat{\mt B}_k} \diff{}{\vt q_i} + \tr{\vt \pi_i}\tr{\hat{\mt B}_k} \diff{}{\vt \pi_i} \right).
\end{split}
\]

Therefore, a symplectic integrator can be devised using a splitting scheme with the action of each propagator (boost, free translation, and free uniaxial rotation) occurring simultaneously for all rigid bodies in the system. If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\sum_{i=1}^N \vt p_i = \vt P$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\sum_{i=1}^N(\vt p_i t - m_i \vt r_i) = \boldsymbol{\mathcal G}$. If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.

By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete free rotation, the Liouville operator corresponding to the original case of Eq.~\eqref{eq:EDO_system} (as best noticed in its alternative form, Eq.~\eqref{eq:EDO_system_alternative}) can be written as
\begin{equation}
\label{eq:full_operator_uniaxial}
i\!L = i\!L_t + i\!L_b + i\!L_r,
\end{equation}
where $i\!L_t$ corresponds to the first term in the right-hand of Eq.~\eqref{eq:tras_rot_op}. The factorization we employ here for the Liouville operator in Eq.~\eqref{eq:full_operator} is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
\end{equation}

Update of forces and torques is required only after the free translation and free rotation have been carried out. Following Miller \textit{et al}.,\cite{Miller2002} each complete rotation is split as
\begin{equation}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} = \prod_{j=1}^n \left( e^{\frac{\delta t}{2} i\!L_3} e^{\frac{\delta t}{2} i\!L_2} e^{\delta t i\!L_1} e^{\frac{\delta t}{2} i\!L_2} e^{\frac{\delta t}{2} i\!L_3} \right),
\end{equation}
where ${\delta t} = {\Delta t}/n$. This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement.
